---
title: "MSDS 6372 Project 1"
author: "Karthik Radhakrishnan"
date: "2024-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis of Hospitalization Stays Data

### Load the data 
```{r}
# Load the data
HospitalStays <- read.csv("HospitalDurations.csv")
head(HospitalStays)

# Find if any of the column has missing values
sum(is.na(HospitalStays))

# Convert Categorical variables into factors

HospitalStays$Region <- as.factor(HospitalStays$Region)
HospitalStays$Med.Sc.Aff <- as.factor(HospitalStays$Med.Sc.Aff)


# Rename columns for better readability
colnames(HospitalStays) <- c("ID", "Length_of_Stay", "Age", "Infection_Risk", "Routine_Culturing_Ratio",
                             "Routine_Chest_Xray_Ratio", "Number_of_Beds", "Medical_School_Affiliation", 
                             "Region", "Avg_Daily_Census", "Number_of_Nurses", "Available_Facilities")


```
### Check Linear Relationships Between Numerical Variables and the Response Variable

```{r}
# Correlation matrix for numerical variables
numerical_cols <- c("Length_of_Stay", "Age", "Infection_Risk", "Routine_Culturing_Ratio", 
                    "Routine_Chest_Xray_Ratio", "Number_of_Beds", "Avg_Daily_Census", 
                    "Number_of_Nurses", "Available_Facilities")


# Calculate correlation matrix
cor_matrix <- cor(HospitalStays[numerical_cols], use = "complete.obs")

# Visualize correlation matrix
library(corrplot)
corrplot(cor_matrix, method = 'color', addCoef.col = "black", number.cex = 0.7)

```

**Infection Risk has positive correlation with Hospital length of stay. Number of Beds and Average daily census have moderate positive correlation with hospital length of stay.**
  
### Scatter plots for numerical variables vs Length of Stay

```{r}

library(ggplot2)

for (col in numerical_cols[-1]) {
  p <- ggplot(data = HospitalStays, aes(x = !!sym(col), y = Length_of_Stay)) +
    geom_point() +
    geom_smooth() +
    labs(title = paste("Scatter Plot of", col, "vs Length_of_Stay"),
         x = col,
         y = "Length of Stay") +
    theme_minimal()
  
  print(p)  # Print the plot object
}
```

**None of the predictors have clear linear relationship with length of Hospital Stay**

### Check Confounding of Categorical Variables with Numerical Variables

```{r}
# Boxplot: Medical School Affiliation vs Numerical Variables
boxplot(Length_of_Stay ~ Medical_School_Affiliation, HospitalStays,
        main="Length of Stay vs Medical School Affiliation", xlab="Medical School Affiliation (1=Yes, 2=No)", ylab="Length of Stay")

# Boxplot: Region vs Length of Stay
boxplot(Length_of_Stay ~ Region, data=HospitalStays,
        main="Length of Stay vs Region", xlab="Region (1=NE, 2=NC, 3=S, 4=W)", ylab="Length of Stay")

```
**Region 1 has higher mean length of Stay compared to other Regions**

### Build a regression Model with all predictors

```{r}
# Create a linear model to predict Length of Stay
model <- lm(Length_of_Stay ~ Age + Infection_Risk + Routine_Culturing_Ratio + Routine_Chest_Xray_Ratio + 
            Number_of_Beds + Medical_School_Affiliation + Region + Avg_Daily_Census + 
            Number_of_Nurses + Available_Facilities, data=HospitalStays)

# Summary of the model
summary(model)

library(car)

Anova(model,type=3)

# Calculate RMSE
residuals <- residuals(model)
rmse <- sqrt(mean(residuals^2))

# Calculate AIC
aic_value <- AIC(model)

# Calculate BIC
bic_value <- BIC(model)

# Print results
cat("RMSE:", rmse, "\nAIC:", aic_value, "\nBIC:", bic_value, "\n")

```

** As per this model without complexity, Infection risk is associated with a patient's length of stay even after accounting for other variables that may impact hospital stays**

### Check for multi colinearity
```{r}
# Load the car package for VIF
library(car)

# Calculate VIF to check for multicollinearity
vif(model)
```

**Number_of_Beds (5.97) and Avg_Daily_Census (5.85) have the highest values, suggesting potential multicollinearity issues with these variables.**

### Check the assumptions of the model built
```{r}
par(mfrow=c(2,2))
plot(model)
par(mfrow=c(1,1))
```


 **Constant Variance assumption is not met in this simple MLR model with all the predictor variables**

### Feature selection using Lasso using glmnet library after Scaling 

``` {r}

# Convert the factor variables into numeric for scaling 

HospitalStays$Region <- as.numeric(HospitalStays$Region)
HospitalStays$Medical_School_Affiliation <- as.numeric(HospitalStays$Medical_School_Affiliation)

# Extracting the predictor variables (excluding the outcome variable, 'Lgth of Sty')
X <- HospitalStays[, c('Age', 'Infection_Risk', 'Routine_Culturing_Ratio', 'Routine_Chest_Xray_Ratio', 'Number_of_Beds', 
              'Medical_School_Affiliation', 'Region', 'Avg_Daily_Census', 'Number_of_Nurses', 'Available_Facilities')]

# Extracting the outcome variable (Length of Stay)
y <- HospitalStays$Length_of_Stay

# Scaling the predictor variables
X_scaled <- scale(X)

# Performing Lasso regression
library(glmnet)
lasso_model <- glmnet(X_scaled, y, alpha = 1)

# Cross-validation to find the optimal lambda 
cv_lasso <- cv.glmnet(X_scaled, y, alpha = 1)

# Get the best lambda value that minimizes cross-validated error
best_lambda <- cv_lasso$lambda.min

# Refit the Lasso model using the best lambda
best_lasso_model <- glmnet(X_scaled, y, alpha = 1, lambda = cv_lasso$lambda.1se)

# Display the coefficients of the final model
coef(best_lasso_model)

# Plot the cross-validation curve
plot(cv_lasso)

#Convert the numerical values into factors again

HospitalStays$Region <- as.factor(HospitalStays$Region)
HospitalStays$Medical_School_Affiliation <- as.factor(HospitalStays$Medical_School_Affiliation)


str(HospitalStays)
```

### Feature selection by Forward selection

```{r}
library(leaps)
reg.fwd=regsubsets(Length_of_Stay~.,data=HospitalStays,method="forward",nvmax=20)
summary(reg.fwd)$bic

par(mfrow=c(1,3))
bics<-summary(reg.fwd)$bic
plot(1:13,bics,type="l",ylab="BIC",xlab="# of predictors")
index<-which(bics==min(bics))
points(index,bics[index],col="red",pch=10)

coef(reg.fwd,7)

```
### Build a model with features selected by lasso regression 

```{r}
# Create a linear model to predict Length of Stay
model2 <- lm(Length_of_Stay ~ Age + Infection_Risk + Region + Routine_Chest_Xray_Ratio+ Avg_Daily_Census, data=HospitalStays)

# Summary of the model
summary(model2)

# Check the assumptioins

par(mfrow=c(2,2))
plot(model2)
par(mfrow=c(1,1))

# Calculate RMSE
residuals <- residuals(model2)
rmse <- sqrt(mean(residuals^2))
real_rmse <- exp(rmse)

# Calculate MAE

# Calculate predictions in the log scale
predictions <- predict(model2)


# Calculate MAE in the original scale
mae_original <- mean(abs(predictions - HospitalStays$Length_of_Stay))


# Print results
cat("RMSE original Scale for MLR Model 0:", real_rmse)
cat("MAE original Scale for MLR model 0:", mae_original)

```


**Constant Variance assumption is not met**


### Log transformation of Y with only the features selected by Lasso regression method 

```{r}
model3 <- lm(log(Length_of_Stay) ~ Age + Infection_Risk + Region + Routine_Chest_Xray_Ratio+ Avg_Daily_Census, data=HospitalStays)

# Summary of the model
summary(model3)

confint(model3)

Anova(model3,type=3)

par(mfrow=c(2,2))
plot(model3)
par(mfrow=c(1,1))

# Calculate RMSE
residuals <- residuals(model3)
rmse <- sqrt(mean(residuals^2))
real_rmse <- exp(rmse)

# Calculate MAE

# Calculate predictions in the log scale
log_predictions <- predict(model3)

# Exponentiate the log predictions to get predictions in the original scale
predictions_original_scale <- exp(log_predictions)

# Calculate MAE in the original scale
mae_original <- mean(abs(predictions_original_scale - HospitalStays$Length_of_Stay))


# Print results
cat("RMSE original Scale for MLR Model 1:", real_rmse)
cat("MAE original Scale for MLR model 1:", mae_original)

```



### MLR model with added complexity

```{r}
# Adding a first-degree polynomial term for Infection_Risk & Interaction with Region

model_poly <- lm(log(Length_of_Stay) ~ Age + poly(Infection_Risk, 1) * Region + Avg_Daily_Census +  Routine_Chest_Xray_Ratio, data = HospitalStays)


# Summary of the model
summary(model_poly)


par(mfrow=c(2,2))
plot(model_poly)
par(mfrow=c(1,1))

# Calculate RMSE
residuals <- residuals(model_poly)
rmse <- sqrt(mean(residuals^2))
real_rmse <- exp(rmse)

# Calculate MAE

# Calculate predictions in the log scale
log_predictions <- predict(model_poly)

# Exponentiate the log predictions to get predictions in the original scale
predictions_original_scale <- exp(log_predictions)

# Calculate MAE in the original scale
mae_original <- mean(abs(predictions_original_scale - HospitalStays$Length_of_Stay))


# Print results
cat("RMSE original Scale for MLR Model 2:", real_rmse)
cat("MAE original Scale for MLR model 2:", mae_original)

```

