---
title: "DS-6372: Project 1 - Objective 2"
author: "Robert Hines"
date: "2024-09-29"
output: html_document
---

```{r message=FALSE}
library(dplyr)         # A Grammar of Data Manipulation and Transformation
library(car)           # Provides tools for regression modeling, including the 'vif' function
library(caret)         # Used for machine learning
library(glmnet)        # For LASSO and Ridge feature selection
library(randomForest)  # For Random Forest implementation

library(ggplot2)       # Elegant Data Visualizations Using the Grammar of Graphics
library(ggcorrplot)    # For creating correlation plots
library(GGally)        # For creating scatter plot matrices
```

```{r}
file_path <- file.choose()
HospitalDurations <- read.csv(file_path)
```

```{r}
colnames(HospitalDurations) <- c("LengthOfStay",    # Lgth of Sty
                                 "ID",              # ID
                                 "Age",             # Age
                                 "InfectionRisk",   # Inf Risk
                                 "CulturesRatio",   # R Cul Rat
                                 "XrayRatio",       # R CX-ray Rat
                                 "NumBeds",         # N Beds
                                 "MedSchoolAff",    # Med Sc Aff
                                 "Region",          # Region
                                 "AvgPatients",     # Avg Pat
                                 "AvgNurses",       # Avg Nur
                                 "PctServices")     # Pct Ser&Fac
```

# Begin Exploratory Data Analysis: View the structure of the dataset
```{r}
str(HospitalDurations)
```

```{r}
summary(HospitalDurations)
```

# Calculating and visualizing the proporiton of missing values in each column of the data: No Missing Values
```{r}
missing_prop <- colMeans(is.na(HospitalDurations))
missing_df <- data.frame(variable = names(HospitalDurations), prop_missing = missing_prop)
missing_df <- missing_df %>% filter(prop_missing != 0)

head(missing_df)
```

# Checking the data for any duplicated values: No Duplicate Values
```{r}
head(HospitalDurations[duplicated(HospitalDurations), ])
```

# Exclude the ID column because it provides no value for analysis
```{r}
HospitalDurations <- HospitalDurations %>%
  dplyr::select(-ID)
```

# Correlation matrix for numeric columns
```{r}
cor_matrix <- cor(HospitalDurations, use = "complete.obs")
cor_matrix
```
#   Plot correlation matrix 
#   Correlations to consider:
#   NumBeds{Avg.Pat:0.98, Avg.Nur:0.92, Med.Sc.Aff:-0.59, Pct.Ser.Fac:0.79}
#   Avg.Pat{NumBeds:0.98, Avg.Nur:0.91, Med.Sc.Aff:-0.51, Pct.Ser.Fac:0.78}
#   Avg.Nur{NumBeds:0.92, Avg.Pat:0.91, Med.Sc.Aff:-0.59, Pct.Ser.Fac:0.78}
#   Pct.Ser.Fac{NumBeds:0.79, Avg.Pat:0.78, Avg.Nur:0.78, Med.Sc.Aff:-0.52}
```{r}
ggcorrplot(cor_matrix, method = "square",  type = "lower",
  lab = TRUE,
  lab_size = 2.5,
  colors = c("#0073C2FF", "white", "#EFC000FF"),
  title = "Correlation Plot of Hospital Durations",
  ggtheme = ggplot2::theme_minimal()
) + 
  theme(legend.position = "none")
```

# Fit a Full Model with all Explanatory Variables
```{r}
model <- lm(LengthOfStay ~ ., data = HospitalDurations)
summary(model)
```

# Check Variance Inflation Factors (VIF) for the model
```{r}
vif_values <- vif(model)
print(vif_values)
```

```{r}
vif_df <- data.frame(variable = names(vif_values), VIF = vif_values)

ggplot(vif_df, aes(x = reorder(variable, VIF), y = VIF)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
  labs(title = "Variance Inflation Factor (VIF) for Each Variable",
       x = "Variable", y = "VIF Value") +
  theme_minimal()
```

# Number of Beds: VIF 34.37 <- This feature is not Statistically Significant. Drop it and retain Average Number of Patients which is highly significant.
```{r}
HospitalDurations <- HospitalDurations %>%
  dplyr::select(-NumBeds)
```

```{r}
model <- lm(LengthOfStay ~ ., data = HospitalDurations)
summary(model)
```

# Check Variance Inflation Factors (VIF) for the model
```{r}
vif_values <- vif(model)
print(vif_values)
```

```{r}
vif_df <- data.frame(variable = names(vif_values), VIF = vif_values)

ggplot(vif_df, aes(x = reorder(variable, VIF), y = VIF)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
  labs(title = "Variance Inflation Factor (VIF) for Each Variable",
       x = "Variable", y = "VIF Value") +
  theme_minimal()
```

```{r}
cor_matrix <- cor(HospitalDurations, use = "complete.obs")
cor_matrix
```

```{r}
ggcorrplot(cor_matrix, method = "square",  type = "lower",
  lab = TRUE,
  lab_size = 2.5,
  colors = c("#0073C2FF", "white", "#EFC000FF"),
  title = "Correlation Plot of Hospital Durations",
  ggtheme = ggplot2::theme_minimal()
) + 
  theme(legend.position = "none")
```

# Create partial regression plots for the model
```{r}
avPlots(model, ask = FALSE) 
```

# Step 1: Prepare the Dataset
```{r}
HospitalDurations <- HospitalDurations %>%
  dplyr::select(-AvgPatients, -AvgNurses) %>%
  mutate(
    Region = as.factor(Region),
    MedSchoolAff = as.factor(MedSchoolAff)
  )
```

```{r}
set.seed(123)
trainIndex <- createDataPartition(HospitalDurations$LengthOfStay, p = 0.8, list = FALSE)
trainData <- HospitalDurations[trainIndex, ]
testData <- HospitalDurations[-trainIndex, ]
```

# 2. LASSO and Ridge Regression
```{r}
x <- model.matrix(LengthOfStay ~ ., trainData)[, -1]
y <- trainData$LengthOfStay
```

```{r}
# Fit LASSO model
lasso_model <- cv.glmnet(x, y, alpha = 1)  # alpha = 1 specifies LASSO
best_lambda_lasso <- lasso_model$lambda.min

# Extract coefficients for best lambda
lasso_coefficients <- as.matrix(coef(lasso_model, s = best_lambda_lasso))
lasso_coeff_df <- data.frame(Feature = rownames(lasso_coefficients), Coefficient = lasso_coefficients[, 1])

# Filter features with non-zero coefficients
lasso_coeff_df <- subset(lasso_coeff_df, Coefficient != 0)
lasso_selected <- rownames(lasso_coefficients)[lasso_coefficients != 0]
```

```{r}
ggplot(lasso_coeff_df, aes(x = reorder(Feature, Coefficient), y = Coefficient)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Non-Zero Feature Coefficients (LASSO Regression)", x = "Features", y = "Coefficient Value")
```

```{r}
# Fit Ridge model
ridge_model <- cv.glmnet(x, y, alpha = 0)  # alpha = 0 specifies Ridge regression
best_lambda_ridge <- ridge_model$lambda.min

# Extract coefficients for best lambda
ridge_coefficients <- as.matrix(coef(ridge_model, s = best_lambda_ridge)) 
ridge_coeff_df <- data.frame(Feature = rownames(ridge_coefficients), Coefficient = ridge_coefficients[, 1])

# Filter features with non-zero coefficients
ridge_coeff_df <- subset(ridge_coeff_df, Coefficient != 0)
ridge_selected <- rownames(ridge_coefficients)[ridge_coefficients != 0]
```

```{r}
ggplot(ridge_coeff_df, aes(x = reorder(Feature, Coefficient), y = Coefficient)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Non-Zero Feature Coefficients (Ridge Regression)", x = "Features", y = "Coefficient Value")
```

```{r}
selected_features <- list(
  LASSO = lasso_selected,
  Ridge = ridge_selected
)
```

# Display the selected features

```{r}
print("Selected Features by LASSO:")
print(selected_features$LASSO)
```

```{r}
print("Selected Features by Ridge:")
print(selected_features$Ridge)
```


```{r}
HospitalDurations <- read.csv(file_path)
```

```{r}
colnames(HospitalDurations) <- c("ID",              # ID
                                 "LengthOfStay",    # Lgth of Sty
                                 "Age",             # Age
                                 "InfectionRisk",   # Inf Risk
                                 "CulturesRatio",   # R Cul Rat
                                 "XrayRatio",       # R CX-ray Rat
                                 "NumBeds",         # N Beds
                                 "MedSchoolAff",    # Med Sc Aff
                                 "Region",          # Region
                                 "AvgPatients",     # Avg Pat
                                 "AvgNurses",       # Avg Nur
                                 "PctServices")     # Pct Ser&Fac
```

```{r}
# Data Preparation
HospitalDurations <- HospitalDurations %>%
  dplyr::select(-ID) %>%
  mutate(
    Region = as.factor(Region),
    MedSchoolAff = as.factor(MedSchoolAff)
  )
```

```{r}
# Define a grid for tuning the 'mtry' parameter
tune_grid <- expand.grid(.mtry = c(2, 3, 4, 5, 6, 7, 8))

# Define different cross-validation methods
cv_methods <- list(
  "5-Fold CV" = trainControl(method = "cv", number = 5),
  "10-Fold CV" = trainControl(method = "cv", number = 10),
  "LOOCV" = trainControl(method = "LOOCV"),
  "Repeated 5-Fold (3x)" = trainControl(method = "repeatedcv", number = 5, repeats = 3),
  "Bootstrap (25 Resamples)" = trainControl(method = "boot", number = 25)
)
```

```{r}
set.seed(123)

# Initialize an empty list to store results and plots
results <- list()
models <- list()
plots <- list()

# Train Random Forest model using each cross-validation method
for (cv_name in names(cv_methods)) {
  cat("Training with", cv_name, "\n")
  rf_model <- train(LengthOfStay ~ ., data = HospitalDurations,
                    method = "rf",
                    trControl = cv_methods[[cv_name]],
                    importance = TRUE,
                    ntree = 1000,
                    metric = "RMSE")

  # Extract metrics (RMSE, MAE, R-squared)
  metrics <- data.frame(
    Method = cv_name,
    RMSE = min(rf_model$results$RMSE),                                     # Minimum RMSE
    MAE = rf_model$results$MAE[which.min(rf_model$results$RMSE)],          # MAE corresponding to best RMSE
    Rsquared = rf_model$results$Rsquared[which.min(rf_model$results$RMSE)] # R-squared for best RMSE
  )
  
  # Store results
  results[[cv_name]] <- metrics
  models[[cv_name]] <- rf_model

  # Store Actual vs Predicted plot
  predictions <- predict(rf_model, HospitalDurations)
  pred_df <- data.frame(Actual = HospitalDurations$LengthOfStay, Predicted = predictions)
  
  plot_actual_vs_predicted <- ggplot(pred_df, aes(x = Predicted, y = Actual)) +
    geom_point(color = "blue", alpha = 0.5) +       
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "solid") +
    labs(title = paste("Actual vs Predicted for", cv_name),
         x = "Predicted Length of Stay",
         y = "Actual Length of Stay") +
    theme_minimal()
  
  # Store the plot in the list
  plots[[paste0(cv_name, "_Actual_vs_Predicted")]] <- plot_actual_vs_predicted

  varImpPlot(rf_model$finalModel, main = paste("Variable Importance for", cv_name))
  
  # Capture the plot to store it
  plots[[paste0(cv_name, "_Variable_Importance")]] <- recordPlot()
}

# Combine results
cv_results <- do.call(rbind, results)
print(cv_results)
```

```{r}
# Identify the best model based on the lowest RMSE
best_cv_method <- cv_results$Method[which.min(cv_results$RMSE)]
best_model <- models[[best_cv_method]]

cat("The best Random Forest model was trained using", best_cv_method, "with an RMSE of", min(cv_results$RMSE), "\n")
print(best_model)
```

```{r}
# RMSE
ggplot(cv_results, aes(x = Method, y = RMSE, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Comparison of RMSE Across Cross-Validation Methods",
       x = "Cross-Validation Method",
       y = "Root Mean Squared Error (RMSE)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# MAE
ggplot(cv_results, aes(x = Method, y = MAE, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Comparison of MAE Across Cross-Validation Methods",
       x = "Cross-Validation Method",
       y = "Mean Absolute Error (MAE)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# R-squared
ggplot(cv_results, aes(x = Method, y = Rsquared, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  labs(title = "Comparison of R-squared Across Cross-Validation Methods",
       x = "Cross-Validation Method",
       y = "R-squared") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Display all stored plots for each cross-validation method
```{r}
for (plot_name in names(plots)) {
  cat("\nDisplaying plot:", plot_name, "\n")
  if (inherits(plots[[plot_name]], "gg")) {
    print(plots[[plot_name]])
  } else {
    replayPlot(plots[[plot_name]])
  }
}
```

# Residual Plotting and Testing Model Assumptions
```{r}
residuals <- HospitalDurations$LengthOfStay - as.numeric(models$`10-Fold CV`$finalModel$predicted)
residual_data <- data.frame(Predicted = as.numeric(models$`10-Fold CV`$finalModel$predicted), Actual = HospitalDurations$LengthOfStay, Residuals = residuals)
```

```{r}
ggplot(residual_data, aes(x = Predicted, y = Residuals)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Predicted Values", x = "Predicted Values", y = "Residuals") +
  theme_minimal()
```

```{r}
ggplot(residual_data, aes(x = Residuals)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Frequency") +
  theme_minimal()
```

```{r}
ggplot(residual_data, aes(sample = Residuals)) +
  stat_qq() +  # Create the Q-Q plot
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Q-Q Plot of Residuals", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```

```{r}
ggplot(residual_data, aes(x = Predicted, y = sqrt(abs(Residuals)))) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "Scale-Location Plot", x = "Predicted Values", y = "√|Residuals|") +
  theme_minimal()
```

#################################################################################################################################################################################################################################################################################################################################################################################################################################################################################                                                                                                                                                                          #########################
#########################                                                            Part 2: Survival Analysis                                                                                     #########################
#########################                                                                                                                                                                          #########################
#########################                                                                                                                                                                          #########################
########################################################################################################################################################################################################################################################################################################################################################################################################################################################

```{r message=FALSE}
library(dplyr)      # A Grammar of Data Manipulation and Transformation
library(ggplot2)    # Elegant Data Visualizations Using the Grammar of Graphics
library(ggcorrplot) # For creating correlation plots
library(car)        # Provides tools for regression modeling, including the 'vif' function
library(caret)      # For data preprocessing

library(survival)   # Used for survival analysis
library(survminer)  # For visualization of survival curves
library(pec)        # Prediction error and concordance measures
library(broom)      # For extracting model coefficients
library(boot)       # For bootstrapping
```

```{r}
file_path <- file.choose()
HospitalDurations <- read.csv(file_path)
```

```{r}
# Data Preparation
HospitalDurations <- HospitalDurations %>%
  dplyr::select(-ID) %>%
  mutate(
    Region = as.factor(Region),
    Med.Sc.Aff = as.factor(Med.Sc.Aff)
  )
```

```{r}
# Rename columns using a more readable and efficient naming convention
colnames(HospitalDurations) <- c("LengthOfStay",    # Lgth of Sty
                                 "Age",             # Age
                                 "InfectionRisk",   # Inf Risk
                                 "CulturesRatio",   # R Cul Rat
                                 "XrayRatio",       # R CX-ray Rat
                                 "NumBeds",         # N Beds
                                 "MedSchoolAff",    # Med Sc Aff
                                 "Region",          # Region
                                 "AvgPatients",     # Avg Pat
                                 "AvgNurses",       # Avg Nur
                                 "PctServices")     # Pct Ser&Fac
```

```{r}
# Calculate Resource to Patient ratio by incorporating the available beds, average patients, and nursing staff
HospitalDurations$ResourceToPatient <- ((HospitalDurations$NumBeds - HospitalDurations$AvgPatients) + HospitalDurations$AvgNurses) / HospitalDurations$AvgPatients
# Why: This measure provides insight into the hospital's resource availability relative to patient load.

# Adjust Infection Risk to a Decimal Scale for consistency
HospitalDurations$InfectionRisk <- HospitalDurations$InfectionRisk / 100
# Why: Converting InfectionRisk to a percentage scale ensures consistency with other risk metrics and allows other feauters to be derived.
```

```{r}
# Calculate PneumoniaRate using the XrayRatio relative to CulturesRatio
HospitalDurations$PneumoniaRate <- HospitalDurations$XrayRatio / (HospitalDurations$CulturesRatio + HospitalDurations$XrayRatio)
# Why: This metric shows how the use of X-rays compares to culture testing for detecting pneumonia. Used for Calculating Pneumonia Rate.

# Calculate the Adjusted Pneumonia Risk by combining infection risk with pneumonia rate
HospitalDurations$PneumoniaRisk <- HospitalDurations$InfectionRisk * HospitalDurations$PneumoniaRate
# Why: Provides an adjusted measure of pneumonia risk by considering both infection and diagnostic rates. Used for Estimated number of Patients Without Pneumonia.
```

```{r}
# Calculate the estimated patients without infection using infection risk
HospitalDurations$EstimatedPatientsWithoutInfection <- HospitalDurations$AvgPatients * (1 - HospitalDurations$InfectionRisk)
# Why: Helps to quantify the number of patients potentially unaffected by infections. Used for Estimated number of Patients Without Pneumonia.
```

```{r}
# Calculate residuals for patients without infection relative to average patients
patients_lm <- lm(EstimatedPatientsWithoutInfection ~ AvgPatients, data = HospitalDurations)
HospitalDurations$PatientsWithoutInfectionResiduals <- residuals(patients_lm)
# Why: Helps identify differences in patient counts that are not explained by the average number of patients.
```

```{r}
HospitalDurations <- HospitalDurations %>%
  dplyr::select(LengthOfStay, Age, InfectionRisk, XrayRatio, MedSchoolAff, Region, AvgPatients, ResourceToPatient, PatientsWithoutInfectionResiduals)
```

```{r}
model <- lm(LengthOfStay ~ Age + InfectionRisk + MedSchoolAff + XrayRatio + AvgPatients + Region*ResourceToPatient + PatientsWithoutInfectionResiduals, data = HospitalDurations)
summary(model)
anova_results <- anova(model)
print(anova_results)
```

```{r}
# Create a histogram for each numeric variable
for (variable in colnames(HospitalDurations)) {
  if (is.numeric(HospitalDurations[[variable]])) {
    p <- ggplot(HospitalDurations, aes_string(x = variable)) +
      geom_histogram(bins = 20, fill = "blue", color = "black", alpha = 0.7) +
      theme_minimal() +
      labs(title = paste("Distribution of", variable), x = variable, y = "Frequency") +
      theme(plot.title = element_text(hjust = 0.5))
    print(p)
  } else {
    # Create a bar plot for categorical variables
    p <- ggplot(HospitalDurations, aes_string(x = variable)) +
      geom_bar(fill = "orange", color = "black", alpha = 0.7) +
      theme_minimal() +
      labs(title = paste("Distribution of", variable), x = variable, y = "Count") +
      theme(plot.title = element_text(hjust = 0.5))
    print(p)
  }
}
```

# Create partial regression plots for the model
```{r}
avPlots(model, ask = FALSE)
```

```{r}
# Log-transforming skewed variables
HospitalDurations$AvgPatients <- log(HospitalDurations$AvgPatients)
HospitalDurations$ResourceToPatient <- log(HospitalDurations$ResourceToPatient)
```

```{r}
# Robust scaling function
robust_scale <- function(x) {
  (x - median(x, na.rm = TRUE)) / IQR(x, na.rm = TRUE)
}

# Apply robust scaling to numeric columns
numeric_columns <- sapply(HospitalDurations, is.numeric)
scaled_predictors <- as.data.frame(lapply(HospitalDurations[, numeric_columns & names(HospitalDurations) != "LengthOfStay"], robust_scale))

# Combine the scaled predictors with non-numeric predictors
non_numeric_predictors <- HospitalDurations[, !numeric_columns]
final_scaled_data <- cbind(scaled_predictors, non_numeric_predictors, LengthOfStay = HospitalDurations$LengthOfStay)
```

```{r}
# Create a histogram for each numeric variable
for (variable in colnames(HospitalDurations)) {
  if (is.numeric(HospitalDurations[[variable]])) {
    p <- ggplot(HospitalDurations, aes_string(x = variable)) +
      geom_histogram(bins = 20, fill = "blue", color = "black", alpha = 0.7) +
      theme_minimal() +
      labs(title = paste("Distribution of", variable), x = variable, y = "Frequency") +
      theme(plot.title = element_text(hjust = 0.5))
    print(p)
  } else {
    # Create a bar plot for categorical variables
    p <- ggplot(HospitalDurations, aes_string(x = variable)) +
      geom_bar(fill = "orange", color = "black", alpha = 0.7) +
      theme_minimal() +
      labs(title = paste("Distribution of", variable), x = variable, y = "Count") +
      theme(plot.title = element_text(hjust = 0.5))
    print(p)
  }
}
```

```{r}
model <- lm(LengthOfStay ~ Age + InfectionRisk + MedSchoolAff + XrayRatio + AvgPatients + Region*ResourceToPatient + PatientsWithoutInfectionResiduals, data = HospitalDurations)
avPlots(model, ask = FALSE)
```


```{r}
# Fit Kaplan-Meier model
km_fit <- survfit(Surv(LengthOfStay) ~ 1, data = HospitalDurations)

# Plot the Kaplan-Meier curve
ggsurvplot(km_fit, conf.int = TRUE)
```

```{r}
# Fit the AFT model using the weibull distribution
aft_weibull <- survreg(Surv(LengthOfStay) ~ Age + InfectionRisk + MedSchoolAff + XrayRatio + AvgPatients + Region*ResourceToPatient + PatientsWithoutInfectionResiduals, 
                           data = HospitalDurations, 
                           dist = "weibull")
summary(aft_weibull)

# Predicted survival times and Performance Metrics
aft_pred <- predict(aft_weibull, type = "response")
aft_rmse <- sqrt(mean((HospitalDurations$LengthOfStay - aft_pred)^2))
cat("RMSE for Weibull AFT Model:", aft_rmse, "\n")
cat("AIC for Weibull AFT Model:", AIC(aft_weibull), "\n")
concordance(Surv(HospitalDurations$LengthOfStay) ~ aft_pred)
```

```{r}
aft_coef <- tidy(aft_weibull)
ggplot(aft_coef, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error)) +
  theme_minimal() +
  labs(title = "Forest Plot of Model Coefficients", x = "Coefficient Estimate", y = "Predictor")
```

```{r}
# Fit the AFT model using the lognormal distribution
aft_lognormal <- survreg(Surv(LengthOfStay) ~ Age + InfectionRisk + MedSchoolAff + XrayRatio + AvgPatients + Region*ResourceToPatient + PatientsWithoutInfectionResiduals, 
                           data = HospitalDurations, 
                           dist = "lognormal")
summary(aft_lognormal)

# Predicted survival times and Performance Metrics
aft_pred <- predict(aft_lognormal, type = "response")
aft_rmse <- sqrt(mean((HospitalDurations$LengthOfStay - aft_pred)^2))
cat("RMSE for LogNormal AFT Model:", aft_rmse, "\n")
cat("AIC for LogNormal AFT Model:", AIC(aft_lognormal), "\n")
concordance(Surv(HospitalDurations$LengthOfStay) ~ aft_pred)
```

```{r}
aft_coef <- tidy(aft_lognormal)
ggplot(aft_coef, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error)) +
  theme_minimal() +
  labs(title = "Forest Plot of Model Coefficients", x = "Coefficient Estimate", y = "Predictor")
```

```{r}
# Fit the AFT model using the log-logistic distribution
aft_loglogistic <- survreg(Surv(LengthOfStay) ~ Age + InfectionRisk + MedSchoolAff + XrayRatio + AvgPatients + Region*ResourceToPatient + PatientsWithoutInfectionResiduals, 
                           data = HospitalDurations, 
                           dist = "loglogistic")
summary(aft_loglogistic)

# Predicted survival times and Performance Metrics
aft_pred <- predict(aft_loglogistic, type = "response")
aft_rmse <- sqrt(mean((HospitalDurations$LengthOfStay - aft_pred)^2))
cat("RMSE for LogLogistic AFT Model:", aft_rmse, "\n")
cat("AIC for LogLogistic AFT Model:", AIC(aft_loglogistic), "\n")
concordance(Surv(HospitalDurations$LengthOfStay) ~ aft_pred)
as.numeric(logLik(aft_loglogistic))
```

```{r}
aft_coef <- tidy(aft_loglogistic)
ggplot(aft_coef, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error)) +
  theme_minimal() +
  labs(title = "Forest Plot of Model Coefficients", x = "Coefficient Estimate", y = "Predictor")
```

```{r}
# Extract residuals from the model
weibull_residuals <- residuals(aft_weibull, type = "deviance")
lognormal_residuals <- residuals(aft_lognormal, type = "deviance")
loglogistic_residuals <- residuals(aft_loglogistic, type = "deviance")
```

# Q-Q Plot & Histogram of Residuals: Check for Normality
```{r}
par(mfrow = c(1, 2))
hist(weibull_residuals, main = "Weibull Deviance Residuals", xlab = "Residuals", col = "lightblue", border = "black")
qqnorm(weibull_residuals, main = "Q-Q Plot of Deviance Residuals")
qqline(weibull_residuals, col = "red")

par(mfrow = c(1, 2))
hist(lognormal_residuals, main = "Lognormal Deviance Residuals", xlab = "Residuals", col = "lightblue", border = "black")
qqnorm(lognormal_residuals, main = "Q-Q Plot of Deviance Residuals")
qqline(lognormal_residuals, col = "red")

par(mfrow = c(1, 2))
hist(loglogistic_residuals, main = "Loglogistic Deviance Residuals", xlab = "Residuals", col = "lightblue", border = "black")
qqnorm(loglogistic_residuals, main = "Q-Q Plot of Deviance Residuals")
qqline(loglogistic_residuals, col = "red")
```

# Residuals vs Fitted Values Plot: Check for Heteroscedasticity
```{r}
fitted_values_weibull <- fitted(aft_weibull)
fitted_values_lognormal <- fitted(aft_lognormal)
fitted_values_loglogistic <- fitted(aft_loglogistic)

ggplot(data = data.frame(Fitted = fitted_values_weibull, Residuals = weibull_residuals), 
       aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  ggtitle("Residuals vs Fitted Values Plot") +
  xlab("Fitted Values") +
  ylab("Residuals") +
  theme_minimal()

ggplot(data = data.frame(Fitted = fitted_values_lognormal, Residuals = lognormal_residuals), 
       aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  ggtitle("Residuals vs Fitted Values Plot") +
  xlab("Fitted Values") +
  ylab("Residuals") +
  theme_minimal()

ggplot(data = data.frame(Fitted = fitted_values_loglogistic, Residuals = loglogistic_residuals), 
       aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  ggtitle("Residuals vs Fitted Values Plot") +
  xlab("Fitted Values") +
  ylab("Residuals") +
  theme_minimal()
```

# Autocorrelation Function Plot: Check for Independence of Residuals
```{r}
acf(weibull_residuals, main = "ACF Plot of Weibull AFT Model Residuals")
acf(lognormal_residuals, main = "ACF Plot of Log-normal AFT Model Residuals")
acf(loglogistic_residuals, main = "ACF Plot of Log-logistic AFT Model Residuals")
```

# Observed vs Predicted Plot: Check Model Fit
```{r}
observed_values <- HospitalDurations$LengthOfStay
predicted_values <- predict(aft_weibull, type = "response")

ggplot(data = data.frame(Observed = observed_values, Predicted = predicted_values), 
       aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  ggtitle("Observed vs Predicted Values") +
  xlab("Observed Length of Stay") +
  ylab("Predicted Length of Stay") +
  theme_minimal()

observed_values <- HospitalDurations$LengthOfStay
predicted_values <- predict(aft_lognormal, type = "response")

ggplot(data = data.frame(Observed = observed_values, Predicted = predicted_values), 
       aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  ggtitle("Observed vs Predicted Values") +
  xlab("Observed Length of Stay") +
  ylab("Predicted Length of Stay") +
  theme_minimal()

observed_values <- HospitalDurations$LengthOfStay
predicted_values <- predict(aft_loglogistic, type = "response")

ggplot(data = data.frame(Observed = observed_values, Predicted = predicted_values), 
       aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  ggtitle("Observed vs Predicted Values") +
  xlab("Observed Length of Stay") +
  ylab("Predicted Length of Stay") +
  theme_minimal()
```

```{r}
# Drop a couple annoying outliers. Not good in real-world use-case.
HospitalDurations <- HospitalDurations %>% filter(row_number() != 47)
HospitalDurations <- HospitalDurations %>% filter(row_number() != 111)
```

```{r}
concordance_index <- function(actual, predicted) {
  concordance <- concordance(Surv(actual) ~ predicted)
  return(concordance$concordance)
}
```

```{r}
# Function for Nested Cross-Validation with Bootstrapping to fit the AFT model and calculate performance metrics
aft_model_metrics <- function(data, indices) {
  
  # Create bootstrapped training data set
  train_data <- data[indices, ]
  
  # Inner loop to perform model selection and fit the AFT model
  model_distributions <- c("weibull", "loglogistic", "lognormal")  # Distributions to evaluate
  inner_results <- list()
  
  for (dist in model_distributions) {
    tryCatch({
      model <- survreg(Surv(LengthOfStay) ~ Age + InfectionRisk + MedSchoolAff + XrayRatio + 
                         AvgPatients + Region * ResourceToPatient + PatientsWithoutInfectionResiduals, 
                       data = train_data, 
                       dist = dist)
      
      # Store results for each model
      inner_results[[dist]] <- list(
        model = model,
        aic = AIC(model),
        logLik = logLik(model)
      )
    }, error = function(e) {
      # Handle cases where the model fails to fit
      inner_results[[dist]] <- list(
        model = NULL,
        aic = Inf,   # Assign a very high AIC value for failed models
        logLik = NA
      )
    })
  }
  
  # Select best model based on AIC
  best_model_name <- names(inner_results)[which.min(sapply(inner_results, function(x) x$aic))]
  best_model <- inner_results[[best_model_name]]$model
  
  # Check if best_model is NULL
  if (is.null(best_model)) {
    return(c(Best_Dist = best_model_name, AIC = NA, LogLik = NA, RMSE = NA, Concordance = NA))
  }
  
  # Use the best model to predict on remaining data
  test_data <- data[-indices, ]
  
  # Ensure we have enough test data to evaluate
  if (nrow(test_data) == 0) {
    return(c(Best_Dist = best_model_name, AIC = NA, LogLik = NA, RMSE = NA, Concordance = NA))  
  }
  
  # Predict survival times using the best model
  predicted <- predict(best_model, newdata = test_data, type = "response")
  
  # Calculate RMSE
  actual <- test_data$LengthOfStay
  rmse <- sqrt(mean((actual - predicted)^2, na.rm = TRUE))
  
  # Calculate Concordance Index (C-index)
  concordance <- concordance_index(actual, predicted) 
  
  # Return performance metrics as a vector
  return(c(
    Best_Dist = best_model_name,
    AIC = AIC(best_model),
    LogLik = as.numeric(logLik(best_model)),
    RMSE = rmse,
    Concordance = concordance
  ))
}
```

```{r}
# Perform bootstrap with 500 resamples
set.seed(123)
bootstrap_results <- boot(data = HospitalDurations, statistic = aft_model_metrics, R = 500)
```

```{r}
results_df <- as.data.frame(bootstrap_results$t)
colnames(results_df) <- c("Best_Dist", "AIC", "LogLik", "RMSE", "Concordance")
```

```{r}
# Summary stats
results_df$AIC <- as.numeric(as.character(results_df$AIC))
results_df$LogLik <- as.numeric(as.character(results_df$LogLik))
results_df$RMSE <- as.numeric(as.character(results_df$RMSE))
results_df$Concordance <- as.numeric(as.character(results_df$Concordance))

summary_metrics <- data.frame(
  Metric = c("AIC", "LogLik", "RMSE", "Concordance"),
  Mean = colMeans(results_df[, 2:5], na.rm = TRUE),
  SD = apply(results_df[, 2:5], 2, sd, na.rm = TRUE),
  Median = apply(results_df[, 2:5], 2, median, na.rm = TRUE),
  IQR = apply(results_df[, 2:5], 2, IQR, na.rm = TRUE)
)

print(summary_metrics)
```

```{r}
results_df$ModelIndex <- 1:nrow(results_df)

# Lowest AIC
best_model_aic <- results_df[which.min(results_df$AIC), ]

# Lowest RMSE
best_model_rmse <- results_df[which.min(results_df$RMSE), ]

# Highest concordance
best_model_concordance <- results_df[which.max(results_df$Concordance), ]

# Display the best models
print(best_model_aic)
print(best_model_rmse)
print(best_model_concordance)
```
